---
title: "Homework 6"
author: "Jarett Smith, Max Thompson, Charlie Deaton, and Balin Allred"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(arules)
library(dplyr)
library(knitr)
TRANS<- read_xlsx("Transactions.xlsx")
TRANS<- TRANS %>%
  group_by(ID) %>%
  summarize(Item=list(unique(Item)))

TRANS<- as(TRANS$Item, 'transactions')
```

# Question 1

1. In the following, we have ten market-basket transactions:

> (a) What is the maximum size of frequent itemsets that can be extracted (assuming minsup = 0)?

**Response: There are a maximum of 5 itemsets.**

> (b) What is the maximum number of association rules that can be extracted from this data set (including rules that have zero support)?

```{r}
(3^5)-(2^6)+1
```

**Response: There are 180 possible rules.**

> (c) Compute the support, confidence, and lift for the association rule {A, D} -> {E}

```{r}
# Support
support<- 4/10
support
```

```{r}
# Confidence
confidence<- 4/4
confidence
```

```{r}
# Lift
lift<- confidence/(6/10)
lift
```

> (d) Find all the frequent itemsets assuming minsup count = 2.

```{r, results='hide'}
rules<- apriori(TRANS, parameter=list(support=0.2, conf = 0))
table<- inspect(rules)
```

```{r}
kable(table)
```

> (e) Find an itemset (of size 2 or larger) that has the largest support.

**Response: Itemset [B, D] and [D, E] has the largest support at 6/10.**

> (f) Find a pair of items, say x and y, such that {x} -> {y} and {y} -> {x} have the same confidence.

**Response: {A} -> {C} and {C} -> {A} have the same confidence of .4**

```{r, results='hide'}
rules<- apriori(TRANS, parameter=list(support=0.2, conf=0, minlen=2))
table<- inspect(subset(rules, lhs %in% c("A", "C")))
```
```{r}
kable(head(table, 2))
```

# Question 2

```{r}
# Given a k-itemset X and an l-itemset Y ⊂ X from a set of transactions T. Given Y is infrequent and l < k − 1. Let Z represent a l + 1 itemset meaning it has 1 more element than Y. Since Z is the itemset Y plus one item, it is a superset of Y. By problem 2 hint, this means that Z is infrequent since Y is infrequent and Z is a superset of Y. Since l < k - 1 is given, we have l + 1 < k, and therefore Z ⊂ X. (Here we are showing there exists itemset l + 1, Z is infrequent and contained in X, so it may be somewhat incorrect. The next section is more of an intuitive proof that may be better.)

# Alternative: We know Y has at least two items less than X since l < k - 1. If we construct Z such that it has one item less than X, it could be represented by itemset k - 1. Then we would have       l < k - 1 < k. Then we know that Y ⊂ Z ⊂ X. Since Y is infrequent we know both Z and X are infrequent. So we have shown there exists an infrequent (k − 1)-itemset Z ⊂ X.
```

